Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-24 main.tex
new version of 45

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Dominic J. D. Hughes}
\IEEEauthorblockA{}
\and
\IEEEauthorblockN{Lutz Straßburger}
\IEEEauthorblockA{Inria, Equipe Partout\\
Ecole Polytechnique, LIX \\ %UMR 7161\\
France}
\and
\IEEEauthorblockN{Jui-Hsuan Wu}
\IEEEauthorblockA{Ecole Normale Supérieure\\
France}}

@@

\todo{Examples,examples,examples}
\section{Introduction}

@@

The main reason for this problem is that formal proofs, as they are
usually studied in logic, are inextricably tied to the syntactic
(inference rule based) proof system in which they are carried out. And
it is difficult to compare two proofs that are produced within two
different syntactic proof systems, based on different sets of inference rules.
\lutz{an example here?}

@@

\begin{equation*}
\vlderivation{
  \vlin{\vlor}{}{\sqn{((\cneg{p} \vlor q) \vlan \cneg{p}) \vlor p}}{
    \vlin{\conr}{}{\sqn{(\cneg{p} \vlor q) \vlan \cneg{p}, p}}{
      \vliin{\vlan}{}{\sqn{(\cneg{p} \vlor q) \vlan \cneg{p}, p, p}}{
        \vlin{\vlor}{}{\sqn{\cneg{p} \vlor q, p}}{
          \vlin{\weakr}{}{\sqn{\cneg{p}, q, p}}{
            \vlin{\axr}{}{\sqn{\cneg{p}, p}}{
              \vlhy{}}}}}{
        \vlin{\axr}{}{\sqn{\cneg{p}, p}}{
          \vlhy{}}}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\cD}{}{((\cneg{p} \vlor q) \vlan \cneg{p}) \vlor p}{
    \vlin{\fequ}{}{((\cneg{p} \vlor q) \vlan \cneg{p}) \vlor (p \vlor p)}{
      \vlin{\sw}{}{(\cneg{p} \vlan (\cneg{p} \vlor q)) \vlor p) \vlor p}{
        \vlin{\fequ}{}{(\cneg{p} \vlan ((\cneg{p} \vlor q) \vlor p)) \vlor p)}{
          \vlin{\sw}{}{(p \vlor (\cneg{p} \vlor q)) \vlan \cneg{p}) \vlor p}{
            \vlin{\aiD}{}{(p \vlor (\cneg{p} \vlor q)) \vlan (\cneg{p} \vlor p)}{
              \vlin{\tttD}{}{(p \vlor (\cneg{p} \vlor q)) \vlan \ttt}{
                \vlin{\fequ}{}{p \vlor (\cneg{p} \vlor q)}{
                  \vlin{\wrD}{}{(p \vlor \cneg{p}) \vlor q}{
                    \vlin{\aiD}{}{p \vlor \cneg{p}}{
                      \vlhy{\ttt}}}}}}}}}}}}
\end{equation*}

\begin{equation*}
\vlderivation{
  \vlin{\conr}{}{\sqn{\exists x.(\cneg{p}x \vlor (\forall y.py))}}{
    \vlin{\exists}{}{\sqn{\exists x.(\cneg{p}x \vlor (\forall y.py)),\exists
x.(\cneg{p}x \vlor (\forall y.py))}}{
      \vlin{\vlor}{}{\sqn{\cneg{p}w \vlor (\forall y.py), \exists
x.(\cneg{p}x \vlor (\forall y.p(y)))}}{
        \vlin{\forall}{}{\sqn{\cneg{p}w, \forall y.py, \exists
x.(\cneg{p}x \vlor (\forall y.py))}}{
          \vlin{\exists}{}{\sqn{\cneg{p}w, pz, \exists
x.(\cneg{p}x \vlor (\forall y.py))}}{
            \vlin{\vlor}{}{\sqn{\cneg{p}w, pz, \cneg{p}z \vlor (\forall y.py))}}{
              \vlin{\weakr}{}{\sqn{\cneg{p}w, pz, \cneg{p}z, (\forall y.py))}}{
                \vlin{\weakr}{}{\sqn{\cneg{p}w, pz, \cneg{p}z}}{
                \vlin{\axr}{}{\sqn{pz, \cneg{p}z}}{
                  \vlhy{}}}}}}}}}}}
\end{equation*}
\juihsuan{LK1 and KS1 proofs of Peirce's law and LK1 proof of drinker formula}
\lutz{can you also do the KS1 for the drinker?}

@@

To give some examples, we show in Figure~\ref{fig:cps} for examples of
FOCPs, each above the formula it proves.  The same FOCPs are shown in
Figure~\ref{fig:cps-condensed} in a more condensed notation.

@@

  $x\in\VAR$. Let us write $\acDx$ for such instances. To address (i),
  we consider here formulas equivalent modulo $\fequ$, always knowing
  that we can add instances of $\fequ$ as needed.\footnote{Note that whenever we have formulas $P$ and $Q$ with $\PE P\PE\fequ\PE Q$ then $P\fequ Q$.} \juihsuan{this does
    not seem clear to me. What if from $\PE A_1$ to $\PE B_1$ there
    are just a bunch of $\PE\fequ$? What do we do in this case?}\lutz{see footnote} To
  address (ii), we apply a permutation argument, permuting all

@@

  \item $\me$ that is not in the situation of~\eqref{eq:mac}: Then now
    encoding of a quantifier is affected and the instance of $\me$ can
    be lifted.\todo{medial permutation!!!}
  \item $\me/\acDx$ as in situation~\eqref{eq:mac}: We must have

@@

    F$. Otherwise, the application of $\acDeq$ would not be correct.
    We have the following four cases: 
    \item $E$ and $F$ are both non-empty: Then~\eqref{eq:mac} is (modulo
      omitted applications of $\fequ$):

@@

    \juihsuan{maybe need some words to exclude the case in which $C$ (or $D$)
is a propositional variable.}\lutz{shit. (you mean a ``first order variable'') this actually can happen. then we have another $\mexD$}
    \item $E$ is empty and $F$ is not: Then~\eqref{eq:mac} becomes
          \vlin{\acDeq}{}{\Scons{(x\vlor F)\vlan(C\vlor D)}}{
            \vlin{\me}{}{\Scons{(x\vlor (x\vlor F))\vlan(C\vlor D)}}{
          \vlhy{\Scons{(x\vlan C)\vlor ((x\vlor F)\vlan D)}}}}}
      The conclusion is the propositional encoding of $\Scons{(\forall
        x. F)\vlan(C\vlor D)}$ and the premise is the propositional
      encoding of $\Scons{(\exists x.C)\vlor ((\forall x. F)\vlor
        D)}$. Also note that no $\me$-instance can break up the
      conjunction in $x\vlan C$ in the premise. Hence, $\phi$ maps an
      existential to a universal, which is ruled out by the
      definition. Hence, this case cannot occur.
    \item $E$ is non-empty and $F$ is empty: This case is similar to the
      previous subcase.
    \item   $E$ and $F$ are both empty: Then~\eqref{eq:mac} is

@@

>>>>

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations

%% \author{\IEEEauthorblockN{Dominic J. D. Hughes}
%% \IEEEauthorblockA{}
%% \and
%% \IEEEauthorblockN{Lutz Straßburger}
%% \IEEEauthorblockA{Inria, Equipe Partout\\
%% Ecole Polytechnique, LIX \\ %UMR 7161\\
%% France}
%% \and
%% \IEEEauthorblockN{Jui-Hsuan Wu}
%% \IEEEauthorblockA{Ecole Normale Supérieure\\
%% France}}

@@

%\todo{Examples,examples,examples}
\section{Introduction}

@@

The main reason for this problem is that formal proofs, as they are
usually studied in logic, are inextricably tied to the syntactic
(inference rule based) proof system in which they are carried out. And
it is difficult to compare two proofs that are produced within two
different syntactic proof systems, based on different sets of inference rules.

\begin{figure}
  $$
  \scalebox{.9}{$
\vlderivation{
  \vlin{\vlor}{}{\sqn{((\cneg{p} \vlor q) \vlan \cneg{p}) \vlor p}}{
    \vlin{\conr}{}{\sqn{(\cneg{p} \vlor q) \vlan \cneg{p}, p}}{
      \vliin{\vlan}{}{\sqn{(\cneg{p} \vlor q) \vlan \cneg{p}, p, p}}{
        \vlin{\vlor}{}{\sqn{\cneg{p} \vlor q, p}}{
          \vlin{\weakr}{}{\sqn{\cneg{p}, q, p}}{
            \vlin{\axr}{}{\sqn{\cneg{p}, p}}{
              \vlhy{}}}}}{
        \vlin{\axr}{}{\sqn{\cneg{p}, p}}{
          \vlhy{}}}}}}
\hskip3em
\vlderivation{
  \vlin{\acD}{}{((\cneg{p} \vlor q) \vlan \cneg{p}) \vlor p}{
    \vlin{\fequ}{}{((\cneg{p} \vlor q) \vlan \cneg{p}) \vlor (p \vlor p)}{
      \vlin{\sw}{}{(\cneg{p} \vlan (\cneg{p} \vlor q)) \vlor p) \vlor p}{
        \vlin{\fequ}{}{(\cneg{p} \vlan ((\cneg{p} \vlor q) \vlor p)) \vlor p)}{
          \vlin{\sw}{}{(p \vlor (\cneg{p} \vlor q)) \vlan \cneg{p}) \vlor p}{
            \vlin{\aiD}{}{(p \vlor (\cneg{p} \vlor q)) \vlan (\cneg{p} \vlor p)}{
              \vlin{\tttD}{}{(p \vlor (\cneg{p} \vlor q)) \vlan \ttt}{
                \vlin{\fequ}{}{p \vlor (\cneg{p} \vlor q)}{
                  \vlin{\wrD}{}{(p \vlor \cneg{p}) \vlor q}{
                    \vlin{\aiD}{}{p \vlor \cneg{p}}{
                      \vlhy{\ttt}}}}}}}}}}}}
$}
  $$
  \caption{Two syntactic proofs of the same formula, in in the sequent calculus (left) and in the calculus of structures (right)}
  \label{fig:exa-synt}
\end{figure}

\lutz{an example here?}

@@

\begin{equation*}
  \scalebox{.7}{$
\vlderivation{
  \vlin{\conr}{}{\sqn{\exists x.(\cneg{p}x \vlor (\forall y.py))}}{
    \vlin{\exists}{}{\sqn{\exists x.(\cneg{p}x \vlor (\forall y.py)),\exists
x.(\cneg{p}x \vlor (\forall y.py))}}{
      \vlin{\vlor}{}{\sqn{\cneg{p}w \vlor (\forall y.py), \exists
x.(\cneg{p}x \vlor (\forall y.p(y)))}}{
        \vlin{\forall}{}{\sqn{\cneg{p}w, \forall y.py, \exists
x.(\cneg{p}x \vlor (\forall y.py))}}{
          \vlin{\exists}{}{\sqn{\cneg{p}w, pz, \exists
x.(\cneg{p}x \vlor (\forall y.py))}}{
            \vlin{\vlor}{}{\sqn{\cneg{p}w, pz, \cneg{p}z \vlor (\forall y.py))}}{
              \vlin{\weakr}{}{\sqn{\cneg{p}w, pz, \cneg{p}z, (\forall y.py))}}{
                \vlin{\weakr}{}{\sqn{\cneg{p}w, pz, \cneg{p}z}}{
                \vlin{\axr}{}{\sqn{pz, \cneg{p}z}}{
                  \vlhy{}}}}}}}}}}}
%\qquad
\vlderivation{
  \vlin{\acD}{}{\exists x.(\cneg{p}x \vlor (\forall y.py))}{
    \vlin{\mfaD}{}{\exists x.(\cneg{p}x \vlor (\forall y.(py \vlor py)))}{
      \vlin{\acD}{}{\exists x.(\cneg{p}x \vlor ((\forall y.py) \vlor (\forall
y.py)))}{
        \vlin{\fequ}{}{\exists x.((\cneg{p}x \vlor \cneg{p}x) \vlor ((\forall
y.py) \vlor (\forall y.py)))}{
          \vlin{\mexD}{}{\exists x.((\cneg{p}x \vlor (\forall y.py)) \vlor
(\cneg{p}x \vlor (\forall y.py)))}{
            \vlin{\exists}{}{(\exists x.(\cneg{p}x \vlor (\forall y.py))) \vlor
(\exists x.(\cneg{p}x \vlor (\forall y.py)))}{
              \vlin{\fequ}{}{(\cneg{p}w \vlor (\forall y.py)) \vlor (\exists
x.(\cneg{p}x \vlor (\forall y.py)))}{
                \vlin{\exists}{}{\forall y.((\cneg{p}w \vlor py) \vlor (\exists
x.(\cneg{p}x \vlor (\forall y.py))))}{
                  \vlin{\fequ}{}{\forall y.((\cneg{p}w \vlor py) \vlor
(\cneg{p}y \vlor (\forall y.py)))}{
                    \vlin{\wD}{}{\forall y.((py \vlor \cneg{p}y) \vlor
(\cneg{p}w \vlor (\forall y.py)))}{
                      \vlin{\aiD}{}{\forall y.(py \vlor \cneg{p}y)}{
                        \vlin{\forall}{}{\forall y.\ttt}{
                          \vlhy{\ttt}}}}}}}}}}}}}}
$}
\end{equation*}

\juihsuan{LK1 and KS1 proofs of Peirce's law and LK1 proof of drinker formula}
\lutz{can you also do the KS1 for the drinker?}
\juihsuan{done}

@@

To give some examples, we show in Figure~\ref{fig:cps} for examples of
FOCPs (taken from~\cite{hughes:fopws}), each above the formula it proves.  The same FOCPs are shown in
Figure~\ref{fig:cps-condensed} in a more condensed notation.

@@

  $x\in\VAR$. Let us write $\acDx$ for such instances.
  The problem with (i) is that we could have the following situation
    \label{eq:m-illegal}
      \vlin{\me}{}{\Scons{((x\vlan E)\vlor(x\vlan F))\vlan(C\vlor D)}}{
        \vlin{\PE\fequ}{}{\Scons{((x\vlan E)\vlan C)\vlor((x\vlan F)\vlan D)}}{
          \vlhy{\Scons{(x\vlan (E\vlan C))\vlor(x\vlan (F\vlan D))}}}}}
  where $x$ occurs in $C\vlor D$. Then premise and conclusion are both
  proposional encodings, but the whole derivation cannot be
  lifted. However, since we demand that the the mapping is a fibration
  (and therefore a momomorphism) on the binding graphs, there must be
  another instance of $\me$ further below in the derivation:
      \vlin{\me}{}{S'\cons{(x\vlor x)\vlan(E\vlor F)}}{
        \vlhy{S'\cons{(x\vlan E)\vlor(x\vlan F)}}}}
  We can permute both instances via the following more general scheme
  (see~\cite{str:medial,lamarche:gap} for a general discussion on
  permutations of the $\me$-rule):
    \label{eq:m-permutation}
    \scalebox{.66}{$
      \vlin{\me}{}{\Scons{(G\vlor H)\vlan(E\vlor F)\vlan(C\vlor D)}}{
        \vlin{\me}{}{\Scons{((G\vlan E)\vlor(H\vlan F))\vlan(C\vlor D)}}{
          \vlhy{\Scons{(G\vlan E\vlan C)\vlor(G\vlan F\vlan D)}}}}}
    \;
    \leftrightarrow
    \;
      \vlin{\me}{}{\Scons{(G\vlor H)\vlan(E\vlor F)\vlan(C\vlor D)}}{
        \vlin{\me}{}{\Scons{(G\vlor H)\vlan((E\vlan C)\vlor(F\vlan D))}}{
          \vlhy{\Scons{(G\vlan E\vlan C)\vlor(G\vlan F\vlan D)}}}}}    
    $}
  We omitted some instances of $\PE\fequ$ and some parentheses.  We
  now call instances of $\me$ as in~\eqref{eq:m-illegal}
  \bfit{illegal}, and we can transform $\PE\Deri_2$ through
  $\me$-permutations~\eqref{eq:m-permutation} into a derivation that
  does not contain any illegal $\me$-instances.
  To
  address (ii),
  we also apply a permutation argument, permuting all

@@

  \item $\me$: There are several cases to consider.
    \item If none of the four principal formulas in the premise is $x$
      or $x\vlor F$ for some formula $F$ and $x\in\VAR$, then this
      instance of $\me$ can trivially be lifted, and and we can
      proceed by induction hypothesis.
  \item If exactly one of the four principal formulas in the premise
    is $x$ for some $x\in\VAR$, then this $x$ is the encoding of an existential in the
    premise and of an universal in the conclusion. This is impossible,
    as $\phi$ has to preserve existentials.
  \item If two of the four principal formulas in the premise
    are $x$ for some $x\in\VAR$, then we are in the following special case of~\eqref{eq:mac}:
  \item We have a situation~\eqref{eq:mac} where

@@

    F$ (Otherwise, the application of $\acDeq$ would not be correct.)
    That means, we have: 

@@

  \item In all other cases (e.g. exactly one of the principal formulas
    is of shape $x\vlor F$ (and none is $x$), we can trivially lift
    the $\me$-instance, as the quantifier structure is not affected.

@@

\hrule
    \begin{itemize}
    \item If none of
    the four principal formulas in the premise is $x$ or $x\vlor F$ or
    $x\vlan F$ for some formula $F$ and $x\in\VAR$, then this instance
    of $\me$ can trivially be lifted, and and we can proceed by
    induction hypothesis.
  \item If exactly one of the four principal formulas in the premise
    is $x$ for some $x\in\VAR$, then this $x$ is the encoding of an existential in the
    premise and of an universal in the conclusion. This is impossible,
    as $\phi$ has to preserve existentials.
  \item If two of the four principal formulas in the premise
    are $x$ for some $x\in\VAR$, then we are in the following special case of~\eqref{eq:mac}:
      \begin{equation*}
        \vlderivation{
          \vlin{\acDeq}{}{\Scons{x\vlan(C\vlor D)}}{
            \vlin{\me}{}{\Scons{(x\vlor x)\vlan(C\vlor D)}}{
              \vlhy{\Scons{(x\vlan C)\vlor (x\vlan D)}}}}}
      \end{equation*}
      which can be lifted immediately to
      \begin{equation*}
        \vlinf{\mexD}{}{\Scons{\exists x.(C\vlor D)}}{\Scons{(\exists x.C)\vlor(\exists x.D)}}
      \end{equation*}

  \item $\me/\acDx$ as in situation~\eqref{eq:mac}: We must have
    $R_1\cons{x}\fequ x\vlor E$ for some $E$ and $R_2\cons{x}\fequ
    x\vlor F$ for some $F$ with $R\cons{x}\fequ x\vlor E\vlor
    F$. Otherwise, the application of $\acDeq$ would not be correct.
    We have the following four cases: 
%    \begin{itemize}
    \item $E$ and $F$ are both non-empty: Then~\eqref{eq:mac} is (modulo
      omitted applications of $\fequ$):
      \begin{equation*}
        \vlderivation{
          \vlin{\acDeq}{}{\Scons{(x\vlor E\vlor F)\vlan(C\vlor D)}}{
            \vlin{\me}{}{\Scons{((x\vlor E)\vlor (x\vlor F))\vlan(C\vlor D)}}{
              \vlhy{\Scons{((x\vlor E)\vlan C)\vlor ((x\vlor F)\vlan D)}}}}}
      \end{equation*}
      which can be lifted to
      \begin{equation*}
        \vlderivation{
          \vlin{\mfaD}{}{\Scons{(\forall x.(E\vlor F))\vlan(C\vlor D)}}{
            \vlin{\me}{}{\Scons{((\forall x. E)\vlor (\forall x. F))\vlan(C\vlor D)}}{
              \vlhy{\Scons{((\forall x. E)\vlan C)\vlor ((\forall x. F)\vlan D)}}}}}
      \end{equation*}
    \juihsuan{maybe need some words to exclude the case in which $C$ (or $D$)
is a propositional variable.}\lutz{shit. (you mean a ``first order variable'') this actually can happen. then we have another $\mexD$}
    \item $E$ is empty and $F$ is not: Then~\eqref{eq:mac} becomes
      \begin{equation*}
        \vlderivation{
          \vlin{\acDeq}{}{\Scons{(x\vlor F)\vlan(C\vlor D)}}{
            \vlin{\me}{}{\Scons{(x\vlor (x\vlor F))\vlan(C\vlor D)}}{
          \vlhy{\Scons{(x\vlan C)\vlor ((x\vlor F)\vlan D)}}}}}
      \end{equation*}
      The conclusion is the propositional encoding of $\Scons{(\forall
        x. F)\vlan(C\vlor D)}$ and the premise is the propositional
      encoding of $\Scons{(\exists x.C)\vlor ((\forall x. F)\vlor
        D)}$. Also note that no $\me$-instance can break up the
      conjunction in $x\vlan C$ in the premise. Hence, $\phi$ maps an
      existential to a universal, which is ruled out by the
      definition. Hence, this case cannot occur.
    \item $E$ is non-empty and $F$ is empty: This case is similar to the
      previous subcase.
    \item   $E$ and $F$ are both empty: Then~\eqref{eq:mac} is
      \begin{equation*}
        \vlderivation{
          \vlin{\acDeq}{}{\Scons{x\vlan(C\vlor D)}}{
            \vlin{\me}{}{\Scons{(x\vlor x)\vlan(C\vlor D)}}{
              \vlhy{\Scons{(x\vlan C)\vlor (x\vlan D)}}}}}
      \end{equation*}
      which can be lifted immediately to
      \begin{equation*}
        \vlinf{\mexD}{}{\Scons{\exists x.(C\vlor D)}}{\Scons{(\exists x.C)\vlor(\exists x.D)}}
      \end{equation*}
    \end{itemize}
